---
title: "Intro to R package development"
author:
    - name: "Lambda Moses"
      email: "dlu2@caltech.edu"
date: "`r Sys.Date()`"
output: 
    BiocStyle::html_document:
        toc: true
        number_sections: true
        toc_depth: 3
        toc_float:
            collapsed: true
vignette: >
  %\VignetteIndexEntry{package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

### Resources
The freely available resources below may be helpful for participating in the hackathon:

* [R Packages](https://r-pkgs.org/): Please read if you don't have prior experiences writing R packages. You will learn about the structure of R packages and how to write documentation and unit tests.
* [Advanced R](https://adv-r.hadley.nz/): This book will give intermediate R users a more in depth understanding of the R language. Please read if you want to tackle issues marked as "advanced", which may involve S4 classes and functional programming.
* [Bioconductor developer guide](https://contributions.bioconductor.org/index.html): Explains in details the requirements for Bioconductor packages. The R Packages book explains CRAN requirements, but Bioconductor has different and often more stringent requirements.
* [Extending `ggplot2`](https://ggplot2.tidyverse.org/articles/extending-ggplot2.html): Advanced `ggplot2` topic relevant to plotting the image behind geometries in `Voyager`.

### System setup
In order to participate in the hackathon, you need to have a [GitHub account](https://github.com/) and [have `git` installed](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git) on your computer. If you haven't used `git` before, read [the Git Basics chapter](https://git-scm.com/book/en/v2/Git-Basics-Getting-a-Git-Repository) to get started. RStudio has a GUI for `git` to make things easier, which is by default in the upper right panel if the R project of interest is version controlled with `git`.

You also need to [install the devel version of R (4.4.0)](https://contributions.bioconductor.org/use-devel.html) and [the RStudio IDE](https://posit.co/download/rstudio-desktop/). The [Xcode Command Line Tools](https://www.freecodecamp.org/news/install-xcode-command-line-tools/) are required for Mac users and [`Rtools`](https://cran.r-project.org/bin/windows/Rtools/) for Windows users if you need to compile dependencies of SFE and Voyager from source, particularly the `sf` package which has several system dependencies. See instructions to install system dependencies of `sf` [here](https://r-spatial.github.io/sf/).

Next `git clone` the repository of interest (Run on the command line (Terminal on Mac and something like Git Bash or PowerShell on Windows), not R console):

```
# SFE
git clone https://github.com/pachterlab/SpatialFeatureExperiment.git

# Voyager
git clone https://github.com/pachterlab/voyager.git
```

This will create a directory `SpatialFeatureExperiment` or `Voyager` where the source code of the package is located. Then in RStudio, [create a project](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects), choose Existing Directory, and RStudio will open the project. 

Next, in RStudio, in the R console, install `devtools` (makes it easier to build, install, and test packages in development), `roxygen2` (to generate documentation pages), `usethis` (makes it easier to add unit tests and vignettes), and `BiocManager` (install and manage Bioconductor packages) from CRAN:

```{r, eval=FALSE}
install.packages(c("devtools", "roxygen2", "usethis", "BiocManager"))
```

Install `devel` version of Bioconductor:

```{r, eval=FALSE}
BiocManager::install(version = "devel")
```

Then install the dependencies; this will also install packages that are suggested but not imported by SFE or Voyager (i.e. soft, optional dependencies) as well as packages required to build and test the package; this can take a while:

```{r, eval=FALSE}
devtools::install_dev_deps()
```

See [Chapter 2 of R Packages](https://r-pkgs.org/setup.html) for more details on setup.

If you work on anything related to `readXenium()` or the `BioFormatsImage` class, please read the instructions to install the optional [`RBioFormats`](https://github.com/aoles/RBioFormats) package because the `BioFormats` Java library is used behind the scene and Java needs special setup on Mac.

## Live demo
We will go over [Chapter 1 of R Packages](https://r-pkgs.org/whole-game.html)

1. Why write a package?
2. Structure of R packages
3. Demo on toy package
4. How to debug
5. Why and how to write unit tests
6. How to do pull requests
7. How to submit packages to Bioconductor (if time permits)

## Voyager vignette guidelines
Vignettes are not in the main or devel branch, but in the `documentation` and `documentation-devel` branches. For the Voyager package, any vignette pull request should fork the `documentation-devel`, which is for vignettes in development. When the new vignette is ready, merge the `documentation-devel` branch into the `documentation` branch, which is the production version of the Voyager documentation website. Rules of the `documentation-devel` branch:

* The purpose of this branch is to make a `pkgdown` website with longer and detailed vignettes that would make the installed size of this package way too large to comply with Bioconductor's 5 MB rule.
* Don't change anything outside the `vignettes` directory in this branch. If the code doesn't work, change in the `main` or `devel` branch and merge into this branch. This way the large vignettes won't get into the `main` branch and the code is kept consistent, which is important since the `pkgdown` website also documents all the functions of this package.
* Exception to the previous rule: you may add packages to the Suggests field in DESCRIPTION for extra packages used in the vignettes.
* The file `vignettes/ref.bib` is automatically synced from Paperpile. Don't edit by hand.
